---
title: "Activity: Decision Trees"
subtitle: "Math 437"
format: pdf
editor: source
execute:
  echo: true
  freeze: false
knitr:
  opts_knit:
    root.dir: "C:\Users\aocho\Math 437 Project Folder\Math 437"
---

## Data

Decision trees can be used for both regression and classification problems. For regression, we'll continue to use our Lahman baseball salary data. For classification, we'll use the RateMyProfessor data.

```{r}
#| label: load packages
library(Lahman)
library(rpart) # don't think we need this, but it's where decision tree algorithms are
library(tidyverse)
library(tidymodels)
```

```{r}
#| label: load batting data
salaries <- Salaries %>%
  dplyr::select(playerID, yearID, teamID, salary)
peopleInfo <- People %>%
  dplyr::select(playerID, birthYear, birthMonth, nameLast,
         nameFirst, bats)
batting <- battingStats() %>% 
  left_join(salaries, 
            by =c("playerID", "yearID", "teamID")) %>%
  left_join(peopleInfo, by = "playerID") %>%
  mutate(age = yearID - birthYear - 
           1L *(birthMonth >= 10)) %>%
  arrange(playerID, yearID, stint)

batting_2016 <- batting %>% filter(yearID == 2016,
                                   !is.na(salary),
                                   G >= 100, AB >= 200
                                   ) %>%
  mutate(salary = log10(salary), # there's a reason for this, I promise
         lgID = factor(lgID, levels = c("AL", "NL"))) # fix the defunct league issue

set.seed(11249)
batting_2016_split <- initial_split(batting_2016, prop = 0.75)
batting_train <- training(batting_2016_split)
batting_test <- testing(batting_2016_split)
```

```{r}
#| label: load RMP data

RMP <- read_csv("../Data Sets/RateMyProfessor.csv")
RMP$hotness <- factor(RMP$hotness, levels = c("hot", "cold"))
RMP$rank <- factor(RMP$rank, 
                   levels = c("Assistant Professor", "Associate Professor", "Professor"),
                   labels = c("Asst", "Assoc", "Full"))
set.seed(1880)
RMP_split <- initial_split(RMP, prop = 0.80) 

RMP_train <- training(RMP_split)
RMP_test <- testing(RMP_split)
```

Decision Trees
Extra high variance models, Data = model +error where the model is highly complex and not much error.
our model"signal" f(x) and error our "noise" where we have high bias models that work well at low SNR(signal to noise ratio) and high variance models that work well at high SNR where main problem is that f_hat(x) is a poor fit to f(x). 





## Regression Trees

1. Explain how a decision tree model works.

When we create a decision tree, there are three parameters that we can tune:

* `tree_depth` indicates the maximum depth of the tree - the default is 30, higher tree depth lower variance
* `min_n` determines whether a node is a terminal node (if there are fewer `min_n` observations at a node, then the node is a terminal node) - the default is 2, higher min_n higher bias
* `cost_complexity` indicates the cost-complexity parameter for tree pruning - the default is 0.01. WE fit the entire decision tree and then "prune", cut out branches that massively increase variance without the correspondingly large decrease in bias

Here we'll leave the first two parameters at their defaults and tune the cost-complexity parameter.

2. Explain the idea behind cost-complexity pruning.

3. Explain the tuning of each parameter in terms of the bias-variance tradeoff.

```{r}
#| label: regression tree model
treeR_model <- decision_tree(
  mode = "regression",
  engine = "rpart", # use rpart to fit the tree
  cost_complexity = tune() # tune this but leave min_n and max_depth at their defaults
)


```

Pre-processing is a good idea but not strictly necessary here.

```{r}
#| label: regression tree recipe
treeR_recipe <- recipe(
  salary ~ G +  R + H + X2B + X3B + HR + RBI + 
    SB + CS + BB + SO + IBB + HBP + SH + SF + GIDP + 
    age + lgID, # response ~ predictors
  data = batting_train
)

treeR_wflow <- workflow() |>
  add_model(treeR_model) |>
  add_recipe(treeR_recipe)
```

Now we can do the cross-validation. We'll use the `grid_regular` function to set up our grid again. If we wanted finer control over our grid, we'd use `expand.grid` to set up our grid manually.

```{r}
#| label: tune regression tree model
set.seed(156)
batting_kfold <- vfold_cv(batting_train,
                          v=5, repeats = 3)

treeR_tune <- tune_grid(
  treeR_wflow,
  resample = batting_kfold,
  grid = grid_regular(
    cost_complexity(range = c(-3,0)
                    ),
    levels = 10
                    ),
  metrics = metric_set(rmse)
  )

```

Just like with penalized regression, the `range` argument is on the log10-scale; in other words, we are searching from $10^{-3}$ (very little pruning) to $10^0 = 1$. 

```{r}
#| label: collect and plot metrics
autoplot(treeR_tune)
```

1. Suppose we are using the one-standard-error rule. Will smaller values of `cost_complexity` or larger values result in simpler models? Why? pick the highest biased model that is still competitive, this makes it a higher biased simpler model. 

2. In my comments I said that we get a warning using R-squared. Recall that we get this warning when every observation is predicted to have the same value. In that case, what would the tree look like?

```{r}
#| label: select best tree
treeR_best <- select_by_one_std_err(
  treeR_tune,
  metric = "rmse",
  desc(cost_complexity)
  
)
treeR_best
```

Now we finalize our workflow and look at how well calibrated the model is:

```{r}
#| label: finalize and fit tree
treeR_wflow_final <- treeR_wflow |>
  finalize_workflow(parameters = treeR_best)
```

```{r}
#| label: investigate calibration of predictions
#calibration: we want the predictions to be on average correct across the entire range of the observed y.
#traditionally for regression: observed y on x axis, predicted y on the y axis(ML doesnt abide by statistical standards)
#we dont look at the test set we use cross validation

treeR_pred_check <- treeR_wflow_final |>
  fit_resamples(
    resamples = batting_kfold,
    control = control_resamples(
      save_pred = TRUE # store the predicted y values
    )
  ) |>
  collect_predictions()  #extract predictions from the output

ggplot(
   treeR_pred_check,
  mapping = aes(x = salary, y = .pred)
) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

```

Well, this certainly looks different from the linear regression models!

3. Explain why we see this pattern - a bunch of horizontal bands - instead.

Assuming we're okay with this somehow-worse-than-linear-regression calibration, we fit the model on the full training set:

```{r}
#| label: fit on full training set
treeR_fit <- fit(treeR_wflow_final,
                 data = batting_train)
```

and investigate what the model actually looks like:

```{r}
#| label: look at tree
par(xpd = TRUE) # makes it so we can see the tree
treeR_fit |>
extract_fit_engine() |>
  plot()
treeR_fit |>
extract_fit_engine() |>
  text(use.n = TRUE)
```

3. Explain how to read this tree.

We then make predictions on the holdout set:

```{r}
#| label: predict with decision tree

```

4. Explain why Javier Baez and Mookie Betts are predicted to have the same salary, even though Baez hit fewer than 14.5 home runs and Betts hit more than 14.5 home runs.

## Classification Trees

Because we already set up the decision tree model, we can just update it to do classification using `set_mode`:

```{r}
#| label: tree model for classification

```

Pre-processing is a good idea but not strictly necessary here. Remember that we are using entirely different data to do entirely different predictions, so we do have to set up the new recipe.

```{r}
#| label: new recipe for classification
treeC_recipe <- recipe(
  hotness ~ scientific_age + gender + 
    interest + uni_control + 
    mentions_accent + discipline + uni_type,  
  data = RMP_train
  )

treeC_wflow <- workflow() |>
  add_model(treeC_model) |>
  add_recipe(treeC_recipe)
```

Now we can do the cross-validation. Traditionally, mean log-loss is used to choose the best decision tree model, so we'll look at that.

```{r}
#| label: tune classification model

```

1. Explain how to read this graph.

```{r}
#| label: select best by 1SE classification

```

Now we finalize our workflow and fit the model on the training set:

```{r}
#| label: fit classification model

```

```{r}
#| label: look at classification tree

```

2. Explain how to read this tree.

```{r}
#| label: augment classification tree

```

### Multiclass Classification

```{r}
#| label: multiclass classification tree
treeMC_recipe <- recipe(
  rank ~ scientific_age + gender + race + uni_control + mentions_accent + discipline + uni_type,  
  data = RMP_train
)

treeMC_wflow <- workflow() |>
  add_model(treeC_model) |>
  add_recipe(treeMC_recipe)
```

```{r}
#| label: tune multiclass tree model

```

```{r}
#| label: select best multiclass tree model

```

```{r}
#| label: fit best multiclass model

```

3. Explain how to read this tree. Why are there so many `scientific_age` splits?
